Evolution of Sex-Specific Recombination Rate

### 2020-05-06 doc/MS_v1.Rmd
fixed annonying typos

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The _Mus musculus_ strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech centre (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken. We use abriviated superscript notation for the remainder of the paper (e.g. _musculus^WSB^_). All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (G) is maintained at UW Veterinary school facilities.

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes based on [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were from mice between 5 and 12 weeks, while juvillne mice ageed between 12 to 15 days were used for DMC1 quantification. Both pairs of ovaries were collected from embryo or neonate mice between 16 to 21 embyronic days and 0 to 48 hours old respctively. Whole testes and decapsulated overaires were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20*C if not stained immediately.

To visualize the strucutre of meiotic chromosomes we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively. The staining protocol was based on that in [@anderson1999] and [@koehler2002]. Antibodies and slide blocking was performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), Triton X-100 (Sigma) ).

@@

An additional possibility is that the different rates in recombination rate variation across subspecies is related to history of hybridization across the populations sampled in this panel of inbred strains. Theoretical work as connected recombination that rate of purging deleterious introgressed DNA in populations after hybridization (Caballero et al 2019, Schumer 2018). The three high-recombining strains identified in this study have connections to hybridization. Both *Musculus^PWD^* and *musculus^SKIVE^* strains were established from samples from the european house mouse hybrid zone. Notably *musculus^SKIVE^* comes from a yonger section of the hybrid zone with a nuclear genome from *M. m. musculus* and a mitochondria genome from *M. m. domesticus* [@gyllensten1987_SKIVE]. The subspecies *M. m. molossinus* is a natural hybrid of *M. m. musculus* and *M. m. cataneus* (Geraldes et al 2008, [@geraldes2011]). Models for the aggregate recombination metics, suggest that the location of recombination events and the total rate influence the dynamics of purging of deleterious introgress DNA (Veller et al 2019), motivating increased study of recombination rates and natural populations.

@@

To place our results in the broader context of recombination rate evolution we compare our main resuts to expectionat under three modifier models and two functional and cell physiology models (table X). While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns for our specific traits across exsisting models. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the chromosome ocillatory movement (COM) model was built to describe positive interference. This spindle based selection model was developed to explain the results resented in this manuscript. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle (bi-orientated) before starting the metaphase-to-anaphase transition [@laneKauppi2019, @subramanian2014, @dumontDesai2012].

@@

Regardless of whether these hypotheses are true, Our observations of between cell variation fail to support a basic prediction of (the relationship between mean and variance) a higher mean will be associated with a higher variance. Across all strains, including the high-recombining strains, _muscuslus^PWD^_, _muscuslus^SKIVE^_ and _molossinus^MSM^_, females consistently have higher between cell variance for crossover count. These results add to previous findings in mice and humans that oogenesis has increased variance for crossovers [@lynn2002, @gruhn2013] and precursors to crossovers [@lenzi2005], compared to spermatocytes. This feature of heterochiasmy supports evolutionary theories predicting distinct reproductive strategies between females (diversifying offspring) and males (maintain successful haplotypes) [@trivers2002].

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers [@schuh2007]. Centrosomes spindles are in spermatogensis and mature sperm cells delivers centrioles (most mammals) or centrosome material (rodents and rabbits) to the egg during fertilization [@manandhar2005centrosome]. This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis [@ross2015].

@@

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (@laneKauppi2019, @subramanian2014, @dumontDesai2012). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate [@vanVeen2003, @altendorfer2020]. Differences in timing of cell cycle between oogenesis and spermatogenesis imposes the the different selective pressures on how sister cohesion affects chromosome structure. Faster spermatogenesis may select for synchronization of the separation homologs (cite). While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle, espicially for those with dicyate arrest [@lee2019].

@@

Our results from _musculus^MSM^_ and _musculus^PWD^_ demonstrate that chromatin compaction are uncoupled from the direction of heterochiasmy in house mice. These results slightly depart from predictions which nominate chromatin compaction as the primary driver of recombination rate variation [@petkov2007]. Chromatin compaction could explain variation in heterochiasmy, which is driven by the disinct meiotic contexts across sexes but is a weak predictor for recombination rate variation within the sexes. Our results indicate the sexual dimorphism in chromatin compaction is conserved in house mouse (e.g. females have longer SC than males) and that the rapid male-specific evolution in crossover number per cell proceeded through another aspect of crossover regulation (see below).

@@

We propose that the spindle based selection model can explain the positive correlation via selection on the amount of sister cohesion connecting homologs at metaphse. The evolution of genome wide recombination rates in our study is driven by a transition of the majority of 
bivalent having single crossovers to an enrichment of bivalents with two crossovers. All else being equal, interference strength would be expected to be equal or even weaker. Our results show that the two crossovers are spaced further apart when SC length is controlled for and chromosome size effects are minimized, in the strain enriched for two crossover bivalents. Assuming that chromatin compaction between pachtyene and metaphase is uniform along chromosomes, an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs (Figure X).

>>>>

To access natural genetic variation for *Mus musculus*, wild derived inbred strains were used. The _Mus musculus_ strains PWD/PhJ, PERC/Eij, WSB/EiJ, LEWE/EiJ, MSM/MsJ, MOLF/EiJ, CAST/EiJ, CZECHII/EiJ were purchased from Jackson labs (Maine, USA). The strains of KAZ/TUA, TOM/TUA, AST/TUA, and HMI/TUA were generated from frozen embryos purchased  from Biological Resource Center (BRC) at Riken (Ibaraki, Japan).  (https://en.brc.riken.jp) and rederived by the UW-Biotech center (Madison UW).
The related murid species *Mus caroli* CAROLI/EiJ and *Mus spretus* SPRET/EiJ were purchased from Jackson labs (https://www.jax.org) and *Mus spicilegus* SPIC/Eij from Riken. We use abbreviated  superscript notation for the remainder of the paper (e.g. _musculus^WSB^_). All mice were housed UW-Madison Biotech and MSC facilities following RARC approved protocols.  A breeding colony of wild derived _Mus musculus_ mice sampled from Gough Island (G) is maintained at UW Veterinary school facilities.

@@

The same dry-down spread technique was applied to both spermatocytes and oocytes based on [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used MLH1 quantification were from mice between 5 and 12 weeks, while juvenile mice aged between 12 to 15 days were used for DMC1 quantification. Both pairs of ovaries were collected from embryo or neonate mice between 16 to 21 embryonic days and 0 to 48 hours old respectively. Whole testes and decapsulated ovaries were incubated in 3ml or 300ul of hypotonic solution for 45min respectively. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20*C if not stained immediately.

To visualize the structure of meiotic chromosomes we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Double strand breaks (DSB) and COs were visualized with DMC1 and MLH1 respectively. The staining protocol was based on that in [@anderson1999] and [@koehler2002]. Antibodies and slide blocking was performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmnuoResearch), 1X PBS, bovine serum albumin (Sigma), Triton X-100 (Sigma) ).

@@

An additional possibility is that the different rates in recombination rate variation across subspecies is related to history of hybridization across the populations sampled in this panel of inbred strains. Theoretical work as connected recombination that rate of purging deleterious introgressed DNA in populations after hybridization (Caballero et al 2019, Schumer 2018). The three high-recombining strains identified in this study have connections to hybridization. Both *Musculus^PWD^* and *musculus^SKIVE^* strains were established from samples from the European house mouse hybrid zone. Notably *musculus^SKIVE^* comes from a younger section of the hybrid zone with a nuclear genome from *M. m. musculus* and a mitochondria genome from *M. m. domesticus* [@gyllensten1987_SKIVE]. The subspecies *M. m. molossinus* is a natural hybrid of *M. m. musculus* and *M. m. cataneus* (Geraldes et al 2008, [@geraldes2011]). Models for the aggregate recombination metrics, suggest that the location of recombination events and the total rate influence the dynamics of purging of deleterious introgress DNA (Veller et al 2019), motivating increased study of recombination rates and natural populations.

@@

To place our results in the broader context of recombination rate evolution we compare our main results to expectations under three modifier models and two functional and cell physiology models (table X). While there are challenges to fitting models built under different assumptions, we feel it's a useful exercise to highlight any unifying patterns for our specific traits across existing models. The three modifier models were built to explain variation in genome wide recombination rates between sexes (heterochiasmy), while the chromosome oscillatory movement (COM) model was built to describe positive interference. This spindle based selection model was developed to explain the results resented in this manuscript. In cases where the models can't be extended to predicting results, we list a 'NA' for the prediction.

@@

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. The SAC prevents aneuploidy by ensuring all bivalents are correctly attached to the microtubule spindle (bi-orientated) before starting the metaphase-to-anaphase transition [@laneKauppi2019, @subramanian2014, @dumontDesai2012].

@@

Regardless of whether these hypotheses are true, our observations of between cell variation fail to support a basic prediction of (the relationship between mean and variance) a higher mean will be associated with a higher variance. Across all strains, including the high-recombining strains, _muscuslus^PWD^_, _muscuslus^SKIVE^_ and _molossinus^MSM^_, females consistently have higher between cell variance for crossover count. These results add to previous findings in mice and humans that oogenesis has increased variance for crossovers [@lynn2002, @gruhn2013] and precursors to crossovers [@lenzi2005], compared to spermatocytes. This feature of heterochiasmy supports evolutionary theories predicting distinct reproductive strategies between females (diversifying offspring) and males (maintain successful haplotypes) [@trivers2002].

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers [@schuh2007]. Centrosomes spindles are in spermatogenesis and mature sperm cells delivers centrioles (most mammals) or centrosome material (rodents and rabbits) to the egg during fertilization [@manandhar2005centrosome]. This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis [@ross2015].

@@

Under the spindle based model we hypothesize that the sexual dimorphism in the recombination landscapes hinges on sex differences on the requirements for chromosome cohesion in late meiosis I. The irreversible process of the metaphase to anaphase transition is initiated by the protyltic decay of the sister cohesion connecting homologs (@laneKauppi2019, @subramanian2014, @dumontDesai2012). The number and placement of crossovers alter the distribution of sister cohesion and the resulting chromosome structure when bivalents are aligned and bi-oriented on the metaphase plate [@vanVeen2003, @altendorfer2020]. Differences in timing of cell cycle between oogenesis and spermatogenesis imposes the the different selective pressures on how sister cohesion affects chromosome structure. Faster spermatogenesis may select for synchronization of the separation homologs (cite). While in oogenesis, the slower cell cycle and multiple arrest stages may require chromosome structures with greater stability on the MI spindle, especially for those with dicyate arrest [@lee2019].

@@

Our results from _musculus^MSM^_ and _musculus^PWD^_ demonstrate that chromatin compaction are uncoupled from the direction of heterochiasmy in house mice. These results slightly depart from predictions which nominate chromatin compaction as the primary driver of recombination rate variation [@petkov2007]. Chromatin compaction could explain variation in heterochiasmy, which is driven by the distinct meiotic contexts across sexes but is a weak predictor for recombination rate variation within the sexes. Our results indicate the sexual dimorphism in chromatin compaction is conserved in house mouse (e.g. females have longer SC than males) and that the rapid male-specific evolution in crossover number per cell proceeded through another aspect of crossover regulation (see below).

@@

We propose that the spindle based selection model can explain the positive correlation via selection on the amount of sister cohesion connecting homologs at metaphase. The evolution of genome wide recombination rates in our study is driven by a transition of the majority of 
bivalent having single crossovers to an enrichment of bivalents with two crossovers. All else being equal, interference strength would be expected to be equal or even weaker. Our results show that the two crossovers are spaced further apart when SC length is controlled for and chromosome size effects are minimized, in the strain enriched for two crossover bivalents. Assuming that chromatin compaction between pachtyene and metaphase is uniform along chromosomes, an outcome of this further spacing of two foci results in an increased area of sister cohesion connecting homologs (Figure X).
